A note on the instability and degeneracy of deep learning models

### 2016-11-20 writing/note.Rmd
remaining edits

---
title: "A note on the instability and degeneracy of deep learning models"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlaated data structures found in several application areas, there is increasing interest in identifying instability in potential model specifications for such data. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes. As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
  Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [@smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. The RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H})$ then has the joint probability mass function
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H}
$$
where 
$$
\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta
$$ 
is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$ are the parameters in the model. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters.

The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
$$
p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N
$$

@@

Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models and an increasing number of parameters. While this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details.

S-unstable FSFS models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following result.

@@

Again, if the probability ratio (\ref{eqn:elpr}) is too large, then the FSFS model will exhibit large probability shifts for very small changes in the data configuration, which exemplifies the intuitive notation of instability..

Additionally, S-unstable FSFS models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space. That is, instability in FSFS models, as measured by (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing levels of one undesirable type of model behavior as (\ref{eqn:elpr}) increases. Furthermore, as (\ref{eqn:elpr}) increases, FSFS models will slide into full degeneracy as Proposition \ref{prop:degen} next shows. Define a modal set 
$$
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$ and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.

@@

In other words, as a consequence of S-unstable models, all probability in the FSFS model formulation with a large number of random variables will stack up on mode sets or simply those few outcomes with the highest probability.

**Proof of Proposition \ref{prop:instab}.** We prove the contrapositive. Suppose that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$. Under the FSFS model, $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$. Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$. Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e. $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$. Then, if $k > 0$, it follows that
\begin{align*}
\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \\
& \le k \Delta_N(\boldsymbol \theta_N) \le NC
\end{align*}
using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$

**Proof of Proposition \ref{prop:degen}.** Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}. Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$. If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (2^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}
\end{align*}
Then, it follows that
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(2^V)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log 2 \rightarrow \infty
\end{align*}
as $N \rightarrow \infty$ by the definition of an unstable FSFS model. Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$

\begin{remark}
There is a further generalization to the notion of "unstable" used in Schweinberger (2011) that deals with replication. One can conceivably imagine replications of the data $\boldsymbol X_1, \dots, \boldsymbol X_n$, where $X_i \in \mathbb{R}^N, i = 1, \dots, n$ such that there are now $n*N$ random variables in the model. However, the definition of S-unstable and propositions \ref{prop:instab}-\ref{prop:degen} still hold if the replications are independent and identically distributed. The reason for this is that for $n$ observations, $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the model and similarly $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest. Thus, 
\begin{align*}
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left\{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right\}^n}{\left\{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right\}^n}\right] \\
&= \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right]\\
&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N).
\end{align*}
Meaning, the definition of an S-unstable FSFS model is invariant to replication.
\end{remark}

@@

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlaated data structures found in several application areas, there is increasing interest in identifying instability in potential model specifications for such data. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes. As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
  Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [@smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. The RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H}
$$
where 
$$
\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta
$$ 
is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$ are the parameters in the model. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters.

The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
$$
p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N
$$

@@

Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g. RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters. While this definition differs in form, it does match that in @schweinberger2011instability for the special exponential models (cf. Section \ref{discrete-exponential-family-models}) considered there.

S-unstable FSFS models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following non-asymptotic result.

@@

Again, if the probability ratio (\ref{eqn:elpr}) is too large, then the FSFS model will exhibit large changes in probability for very small differences in the data configuration, which exemplifies the intuitive notation of instability.

Additionally, S-unstable FSFS models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space. For perspective, note that instability in FSFS models, as measured by (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity to model probabilities (i.e., in one component changes to outcomes) as (\ref{eqn:elpr}) increases. Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS models are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} next shows. Define a $\epsilon$-modal set 
$$
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$, and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.

@@

In other words, in S-unstable FSFS models, all probability in the FSFS model formulation with a large number of random variables will mass on mode sets of arbitrarily small size, or simply those few outcomes with the highest probability.

\begin{remark}
There is a further generalization to the notion of "unstable" used in Schweinberger (2011) that deals with replication. One can conceivably imagine $n$ independent and identically distributed (iid) replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}<\infty|$, for $i=1,\ldots,n$.  This leads to a total of $n*N$ random variables in the joint FSFS model. However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications. The reason for this is that $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability. Thus, 
\begin{align*}
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left\{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right\}^n}{\left\{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right\}^n}\right] \\
&= \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right]\\
&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
\end{align*}
implying that the definition of an S-unstable FSFS model is invariant to the level $n$ of independent replication.
\end{remark}

@@

\begin{appendix}
\section{Appendices}
\label{appendix}
\subsection{Proofs of main results}
\label{proofs}

```{r child='appendix/appendix-note.Rmd'}
```
\end{appendix}
